Spring haul-out behavior of seals in the Bering and Chukchi Seas: implications for abundance estimation

### 2024-09-22 analysis/paper/London_HauloutBehavior.tex
final edits for the proofs

\begin{abstract}
Ice-associated seals rely on sea ice for a variety of activities, including pupping, breeding, molting, and resting. In the Arctic, many of these activities occur in spring (April -- June) as sea ice begins to melt and retreat northward. Rapid acceleration of climate change in Arctic ecosystems is therefore of concern as the quantity and quality of suitable habitat is forecast to decrease. Robust estimates of seal population abundance are needed to properly monitor the impacts of these changes over time. Aerial surveys of seals on ice are an efficient method for counting seals but must be paired with estimates of the proportion of seals out of the water to derive population abundance. In this paper, we use hourly percent-dry data from satellite-linked bio-loggers deployed between 2005 and 2021 to quantify the proportion of seals hauled out on ice. This information is needed to accurately estimate abundance from aerial survey counts of ice-associated seals (i.e., to correct for the proportion of animals that are in the water while surveys are conducted). In addition to providing essential data for survey `availability' calculations, our analysis also provides insights into the seasonal timing and environmental factors affecting haul-out behavior by ice-associated seals. We specifically focused on bearded (\emph{Erignathus barbatus}), ribbon (\emph{Histriophoca fasciata}), and spotted seals (\emph{Phoca largha}) in the Bering and Chukchi seas. Because ringed seals (\emph{Phoca (pusa) hispida}) can be out of the water but hidden from view in snow lairs analysis of their `availability' to surveys requires special consideration; therefore,they were not included in this analysis. Using generalized linear mixed pseudo-models to properly account for temporal autocorrelation, we fit models with covariates of interest (e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature, precipitation) to examine their ability to explain variation in haul-out probability. We found evidence for strong diel and within-season patterns in haul-out behavior, as well as strong weather effects (particularly wind and temperature). In general, seals were more likely to haul out on ice in the middle of the day and when wind speed was low and temperatures were higher. Haul-out probability increased through March and April, peaking in May and early June before declining again. The timing and frequency of haul-out events also varied based on species and age-sex class. For ribbon and spotted seals, models with year effects were highly supported, indicating that the timing and magnitude of haul-out behavior varied among years. However, we did not find broad evidence that haul-out timing was linked to annual sea-ice extent. Our analysis emphasizes the importance of accounting for seasonal and temporal variation in haul-out behavior, as well as associated environmental covariates, when interpreting the number of seals counted in aerial surveys.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

Knowledge of haul-out patterns is not only important for understanding natural
history and ecology, but also for developing ``availability'' correction factors
for aerial surveys. Specifically, researchers need to know the fraction of seals
hauled out (versus in the water) when aerial surveys are conducted. Studies
estimating availability correction factors for seals typically use logistic
regression-style analyses to estimate the time-specific probability of being
hauled out based on `wet/dry' data relayed by bio-loggers. In these models,
haul-out probabilities were expressed as a function of predictive covariates,
such as time-of-day, day-of-year, sex, age class, and environmental conditions
(e.g., Reder et al. (2003), Bengtson \& Cameron (2004), Bengtson et al. (2005), Udevitz et al. (2009), Lonergan et al. (2013),
Ver Hoef et al. (2014), Southwell et al. (2008), and Niemi et al. (2023)). However, sample sizes have often been
insufficient to permit strong inference about demographic and/or seasonal
variation in haul-out probabilities. For instance, Bengtson and Cameron's
(2004) study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s (2005) study was based on 6 ringed seals in the
Chukchi and Beaufort seas. These studies were often further limited by
logistical constraints on fieldwork and the attachment duration or operational
life of bio-loggers. In this study, we addressed some of these limitations by
deploying small bio-loggers designed for longer-term attachment on rear flippers
of a subset of the study individuals. These devices are designed to collect data
through the molt period (when those adhered to the hair -- a more conventional
method -- would fall off) and, in some situations, provide multiple years of
data.

In this study, we used data collected from multiple bio-logging deployments
spanning a 16-year period to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we sought to refine
estimates of haul-out availability corrections for aerial surveys in order to
improve estimates of seal abundance. Previously estimated availability
correction factors (e.g., Bengtson et al. (2005), Conn et al. (2014), and Ver Hoef et al. (2014))
accounted for variables such as the time-of-day and day-of-year, but did not
investigate the impact of weather variables. Such variables have been
shown to influence walrus haul-out behavior (Udevitz et al., 2009) and we expect weather
conditions to also influence seal haul-out behavior and including them within
the model framework will benefit our estimates of seal availability during
aerial surveys. Third, we aimed to assess the annual variability in haul-out
timing and possible linkage to changes in the extent of seasonal sea ice between
2005 and 2021. Our work extends the scope of previous haul-out analyses,
includes the influence of weather variability, and investigates the
potential impact of changing sea-ice extent on the behavior of these species.

@@

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2021. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships (Figure \ref{fig:deployplot}). Ship-based capture events
occurred during spring near the southern ice edge in the Bering Sea between 2005
and 2018. Land-based capture events occurred between 2005 and 2020 from May to
October, generally between the Alaska coastal communities of Scammon Bay in the
Bering Sea, Utqiaġvik in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. Data
from additional deployments along the Kamchatka peninsula in the western Bering
Sea are also included. We refer readers to the primary literature for detailed
capture and bio-logger attachment methods (see publications listed in
Supplemental Material, \hyperref[s1]{S1}). NOAA-led research was conducted under the
authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
Fisheries Science Center Institutional Animal Care and Use Committee (IACUC).
ADF\&G and NSB field work was covered by Research Permits Nos. 358-1585,
358-1787, 15324, and 20466 and by ADF\&G IACUC permits Nos. 06-16, 09-21,
2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.

@@

A typical model fitting exercise would also include a model selection process.
However, AIC (and similar criteria) is not suitable when using
pseudo-likelihoods, because pseudo-data generated in the model fitting process
(Ver Hoef, London \& Boveng, 2010) differ between models (Ten Eyck \& Cavanaugh, 2018). After fitting GLMPM
models, we instead used ``type III'' \(F\)-tests to calculate \(p\)-values
(Ver Hoef, London \& Boveng, 2010) to evaluate model performance and important terms. We also
produced predictions of haul-out behavior as a function of three influential
predictors (e.g.~solar hour, day-of-year, age-sex). Weather covariates for these
predictions were based on daily or hourly smoothed weather covariate values
across the study region. Such predictions were then used to develop haul-out
probability surfaces, explore conditional effects of weather covariates, and
determine annual peaks in haul-out activity. The timing of peak haul-out
behavior was further used to regress against the annual maximum sea-ice extent
in the study region. Predictions before 15 March and after 30 June were not
included in visualizations or other evaluations to avoid spurious model
predictions at the edge of the data range.

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \ref{fig:beardedHOCal}; see also
\ref{fig:beardedPredSE}). Additionally, after approximately 9 months, 7 devices
deployed on the rear flipper of bearded seals reported implausible hourly
percent-dry data (100\% dry for several weeks but indicative of movement and
increasing transmission rates (see Boveng \& Cameron (2013))). All data after the first
instance of unrealistic values were censored from this analysis. In addition to
a peak around local solar noon, the bearded seal model predicted additional
haul-out activity around local midnight. In concert with the lower magnitude of
haul-out probability, bearded seal haul-out behavior was also more protracted
throughout the spring season compared to ribbon
and spotted seals (see below). Overall, bearded seals were
less likely to haul out and had a bi-modal distribution of haul-out probability
across the day.

@@

Ribbon seals exhibited a pattern of gradually increasing haul-out probability in
April that peaks in late May for subadults and in early June for adults (Figure
\ref{fig:ribbonHOCal}; see also \ref{fig:ribbonPredSE}). There is an apparent
seasonal progression with subadults hauling out earlier in the season followed
by adult males and, then, adult females. Haul-out behavior was clearly centered
around local solar noon and expanded to other hours later in the spring as seals
entered their molting period. Subadults showed an earlier start and more intense
haul-out activity in April and May. The young-of-the-year records began after
weaning and the model predictions seemed to indicate development of in-water
activities (e.g.~swimming, foraging) in May and, like adults, haul-out behavior
was centered around solar noon. Adult females had a more protracted
haul-out season compared to males, and more time was spent hauled out in June
compared to adult males and subadults.

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season
that was less intensely centered on solar noon (Figure \ref{fig:spottedHOCal};
see also \ref{fig:spottedPredSE}). Adults of both sexes spent considerable time
in April, May and June hauled out. Adult spotted seal males had a more
protracted haul-out season compared to females, and more time out of the water
in June (Figure \ref{fig:spottedHOCal}). This likely reflects the triad
behavior in spotted seals when suitor males haul out with nursing females. As
with ribbon seals, the young-of-the-year records began after weaning and the
model predictions reflected development of in-water activities (e.g.
swimming, foraging) in May and haul-out behavior centered around solar noon.

@@

The second set of models, which included annual variation in haul-out
patterns, uncovered significant contributions for linear and quadratic
interactions between day and year for only spotted seals (day:year,
\(F_{15,115144}\)
= 4.445; \(p\) =
\textless0.001; day\textsuperscript{2}:year,
\(F_{15,115144}\)
= 5.854; \(p\) =
\textless0.001). Ribbon seals showed no significant
contribution for interactions between day and year (day:year,
\(F_{10,99510}\)
= 0.516; \(p\) =
0.880; day\textsuperscript{2}:year,
\(F_{10,99510}\)
= 0.549; \(p\) =
0.856). Predicted distributions of
haul-out activity were largely unimodal, but varied some among and
within years with respect to the timing and magnitude of haul-out peaks
(Figure \ref{fig:annualHO}). It is important to note that predicted
variation in annual haul-out patterns likely reflected both process error
and sampling variability. While we did remove any years where only one
deployment in a species + age:sex group was present, there were still
some years where the pattern shown was informed by a small number of
individuals that may not represent population-level patterns.

@@

The annual timing of peak haul-out probability for ribbon seals and adult
male spotted seals appeared to have
no relationship with the amount of yearly maximum sea-ice extent in the
study area. For ribbon seals and adult male spotted seals,
\(p\)-values were substantially larger than 0.05 (ribbon seal adult
females; \(R^{2}\) = 0.004, \(p\) = 0.896;
ribbon seal adult males: \(R^{2}\) = 0.059, \(p\) =
0.693; ribbon seal subadults: \(R^{2}\) =
0.007, \(p\) = 0.828; spotted seals
adult males: \(R^{2}\) = 0.004, \(p\) =
0.889). Adult female and subadult spotted seals both
showed a negative trend (peak haul-out occurs later in years with less
sea ice) with a significant relationship for adult female spotted seals
(spotted seal adult female: \(R^{2}\) =
0.456, \(p\) = 0.141; spotted seal
subadults: \(R^{2}\) = 0.369, \(p\) =
0.062).

@@

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g.~Udevitz et al.
(2009)) and a few select studies of ice-associated seals
(Perry, Stenson \& Buren, 2017; Hamilton \& Evans, 2018). In our study, ribbon seal haul-out behavior seemed to be the most
influenced by weather, with wind, temperature, and barometric pressure all being
important components of the model. Spotted seals were most affected by wind and
barometric pressure. For bearded seals, the model indicated wind and temperature
had the greatest impact. In general, and as might be expected, seals were more
likely to haul-out when daily temperatures were warmer, winds speeds were lower,
barometric pressure was higher, and precipitation was lower. Those weather
conditions are general indicators of increased solar radiation and lower
convective heat loss, both of which provide energetic benefits (see additional
discussion in Supplemental Material \hyperref[exploring-insolation-solar-radiation-as-a-model-covariate]{Exploring Insolation (Solar Radiation) as a
Model Covariate} regarding the potential use of solar radiation directly). Low
winds and precipitation could also enhance predator detection. Our results
highlight the importance of incorporating weather covariates when analyzing
haul-out behavior and calculating availability corrections for aerial surveys.

@@

We recognize that the species and ecosystems we studied are within the ancestral
and present-day environs of the Inpuiat and Yup'ik people who, through many
uncredited contributions of traditional knowledge, provided early western
naturalists and scientists with much of what gets described as the `basic
biology' of Arctic seals. The deployment of bio-logging devices used in this
study were often done in collaboration with Alaska Native seal hunters and the
approval of
their communities. We would like to especially acknowledge the communities of
Kotzebue, Koyuk, Nome, Nuiqsut, Scammon Bay, St.~Michael, Utqiaġvik, and Ulguniq
(Wainwright) and the following individuals: James Adams, Jeff Barger, David
Barr, Wendell Booth, Cyrus Harris, Nereus `Doc' Harris, Grover Harris, Lee
Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin,
Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr.,
Virgil Naylor Sr., Dan Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone,
and Randy Toshavik from Kotzebue; Merlin Henry from Koyuk; Tom
Gray from Nome; Vernon Long and Richard Tukle from Nuiqsuit; Morgan
Simon, River Simon, and Al Smith from Scammon Bay; Alex Niksik Jr.~from
St.~Michael; Billy Adams, James Aiken, Tim Aiken, Howard Kittick,
Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, and Joe Skin from Utqiaġvik,
Alaska; Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
Fred Rexford from Ulguniq.

We are grateful for the assistance in catching and sampling seals by Ryan Adam,
James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov,
Cynthia Christman, Sarah Coburn, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey Goldstein, Jeff
Harris, Jason Herreman, Markus Horning, John Jansen, Shawn Johnson, Charles
Littnan, Lloyd Lowry, Brett McClintock, Erin Moreland, Mark Nelson, Justin
Olnes, Lorrie Rea, Bob Shears, Gay Sheffield, Brent Stewart, Dave Withrow, and
Heather Ziel. We also appreciate the commitment to science and safety by all
officers and crew of the NOAA ship \emph{Oscar Dyson}, the NOAA ship \emph{MacArthur II}, and
the RV \emph{Thomas G. Thompson}.

Telemetry data from the Alaska Department of Fish and Game (ADF\&G) and the North
Slope Borough Department of Wildlife Management (NSB) were important
contributions to the findings presented here. Deployments in the western Bering
Sea were done in collaboration with Russian colleagues and North Pacific
Wildlife.

>>>>

\begin{abstract}
Ice-associated seals rely on sea ice for a variety of activities, including pupping, breeding, molting, and resting. In the Arctic, many of these activities occur in spring (April through June) as sea ice begins to melt and retreat northward. Rapid acceleration of climate change in Arctic ecosystems is therefore of concern as the quantity and quality of suitable habitat is forecast to decrease. Robust estimates of seal population abundance are needed to properly monitor the impacts of these changes over time. Aerial surveys of seals on ice are an efficient method for counting seals but must be paired with estimates of the proportion of seals out of the water to derive population abundance. In this paper, we use hourly percent-dry data from satellite-linked bio-loggers deployed between 2005 and 2021 to quantify the proportion of seals hauled out on ice. This information is needed to accurately estimate abundance from aerial survey counts of ice-associated seals (i.e., to correct for the proportion of animals that are in the water, and so are not counted, while surveys are conducted). In addition to providing essential data for survey `availability' calculations, our analysis also provides insights into the seasonal timing and environmental factors affecting haul-out behavior by ice-associated seals. We specifically focused on bearded (\emph{Erignathus barbatus}), ribbon (\emph{Histriophoca fasciata}), and spotted seals (\emph{Phoca largha}) in the Bering and Chukchi seas. Because ringed seals (\emph{Phoca (pusa) hispida}) can be out of the water but hidden from view in snow lairs analysis of their `availability' to surveys requires special consideration; therefore, they were not included in this analysis. Using generalized linear mixed pseudo-models to properly account for temporal autocorrelation, we fit models with covariates of interest (e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature, precipitation) to examine their ability to explain variation in haul-out probability. We found evidence for strong diel and within-season patterns in haul-out behavior, as well as strong weather effects (particularly wind and temperature). In general, seals were more likely to haul out on ice in the middle of the day and when wind speed was low and temperatures were higher. Haul-out probability increased through March and April, peaking in May and early June before declining again. The timing and frequency of haul-out events also varied based on species and age and sex class. For ribbon and spotted seals, models with year effects were highly supported, indicating that the timing and magnitude of haul-out behavior varied among years. However, we did not find broad evidence that haul-out timing was linked to annual sea-ice extent. Our analysis emphasizes the importance of accounting for seasonal and temporal variation in haul-out behavior, as well as associated environmental covariates, when interpreting the number of seals counted in aerial surveys.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

Knowledge of haul-out patterns is not only important for understanding natural
history and ecology, but also for developing `availability' correction factors
for aerial surveys. Specifically, researchers need to know the fraction of seals
hauled out (versus in the water) when aerial surveys are conducted. Studies
estimating availability correction factors for seals typically use logistic
regression-style analyses to estimate the time-specific probability of being
hauled out based on `wet/dry' data relayed by bio-loggers. In these models,
haul-out probabilities were expressed as a function of predictive covariates,
such as time-of-day, day-of-year, sex, age class, and environmental conditions
(e.g., Reder et al. (2003), Bengtson \& Cameron (2004), Bengtson et al. (2005), Udevitz et al. (2009), Lonergan et al. (2013),
Ver Hoef et al. (2014), Southwell et al. (2008), and Niemi et al. (2023)). However, sample sizes have often been
insufficient to permit strong inference about demographic and/or seasonal
variation in haul-out probabilities. For instance, Bengtson and Cameron's
(2004) study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s (2005) study was based on 6 ringed seals in the
Chukchi and Beaufort seas. These studies were often further limited by
logistical constraints on fieldwork and the attachment duration or operational
life of bio-loggers. In this study, we addressed some of these limitations by
deploying small bio-loggers designed for longer-term attachment on rear flippers
of a subset of the study individuals. These devices are designed to collect data
through the molt period (when those adhered to the hair -- a more conventional
method -- would fall off) and, in some situations, provide multiple years of
data.

In this study, we used data collected from multiple bio-logging deployments
spanning a 16-year period to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we sought to refine
estimates of haul-out availability corrections for aerial surveys in order to
improve estimates of seal abundance. Previously estimated availability
correction factors (e.g., Bengtson et al. (2005), Conn et al. (2014), and Ver Hoef et al. (2014))
accounted for variables such as the time-of-day and day-of-year, but did not
investigate the impact of weather variables. Such variables have been
shown to influence walrus haul-out behavior (Udevitz et al., 2009) and we expected weather
conditions to also influence seal haul-out behavior and including them within
the model framework will benefit our estimates of seal availability during
aerial surveys. Third, we aimed to assess the annual variability in haul-out
timing and possible linkage to changes in the extent of seasonal sea ice between
2005 and 2021. Our work extends the scope of previous haul-out analyses,
includes the influence of weather variability, and investigates the
potential impact of changing sea-ice extent on the behavior of these species.

@@

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2021. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships (Figure \ref{fig:deployplot}). Ship-based capture events
occurred during spring near the southern ice edge in the Bering Sea between 2005
and 2018. Land-based capture events occurred between 2005 and 2020 from May to
October, generally between the Alaska coastal communities of Scammon Bay in the
Bering Sea, Utqiaġvik in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. Data
from additional deployments along the Kamchatka peninsula in the western Bering
Sea are also included. We refer readers to the primary literature for detailed
capture and bio-logger attachment methods (see publications listed in
Supplemental Information Table \hyperref[s1]{S1}). NOAA-led research was conducted under the
authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
Fisheries Science Center Institutional Animal Care and Use Committee (IACUC).
ADF\&G and NSB field work was covered by Research Permits Nos. 358-1585,
358-1787, 15324, and 20466 and by ADF\&G IACUC permits Nos. 06-16, 09-21,
2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.

@@

A typical model fitting exercise would also include a model selection process.
However, AIC (and similar criteria) is not suitable when using
pseudo-likelihoods, because pseudo-data generated in the model fitting process
(Ver Hoef, London \& Boveng, 2010) differ between models (Ten Eyck \& Cavanaugh, 2018). After fitting GLMPM
models, we instead used `type III' \(F\)-tests to calculate \(p\)-values
(Ver Hoef, London \& Boveng, 2010) to evaluate model performance and important terms. We also
produced predictions of haul-out behavior as a function of three influential
predictors (e.g.~solar hour, day-of-year, age-sex). Weather covariates for these
predictions were based on daily or hourly smoothed weather covariate values
across the study region. Such predictions were then used to develop haul-out
probability surfaces, explore conditional effects of weather covariates, and
determine annual peaks in haul-out activity. The timing of peak haul-out
behavior was further used to regress against the annual maximum sea-ice extent
in the study region. Predictions before 15 March and after 30 June were not
included in visualizations or other evaluations to avoid spurious model
predictions at the edge of the data range.

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \ref{fig:beardedHOCal}; see also Supplemental Information Fig.
\ref{fig:beardedPredSE}). Additionally, after approximately 9 months, 7 devices
deployed on the rear flipper of bearded seals reported implausible hourly
percent-dry data (100\% dry for several weeks but indicative of movement and
increasing transmission rates (see Boveng \& Cameron (2013))). All data after the first
instance of unrealistic values were censored from this analysis. In addition to
a peak around local solar noon, the bearded seal model predicted additional
haul-out activity around local midnight. In concert with the lower magnitude of
haul-out probability, bearded seal haul-out behavior was also more protracted
throughout the spring season compared to ribbon
and spotted seals (see below). Overall, bearded seals were
less likely to haul out and had a bi-modal distribution of haul-out probability
across the day.

@@

Ribbon seals exhibited a pattern of gradually increasing haul-out probability in
April that peaks in late May for subadults and in early June for adults (Figure
\ref{fig:ribbonHOCal}; see also Supplemental Information Fig.
\ref{fig:ribbonPredSE}). There is an apparent seasonal progression with
subadults hauling out earlier in the season followed by adult males and, then,
adult females. Haul-out behavior was clearly centered around local solar noon
and expanded to other hours later in the spring as seals entered their molting
period. Subadults showed an earlier start and more intense haul-out activity in
April and May. The young-of-the-year records began after weaning and the model
predictions seemed to indicate development of in-water activities (e.g.
swimming, foraging) in May and, like adults, haul-out behavior was centered
around solar noon. Adult females had a more protracted haul-out season compared
to males, and more time was spent hauled out in June compared to adult males and
subadults.

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season
that was less intensely centered on solar noon (Figure \ref{fig:spottedHOCal};
see also Supplemental Information Fig.\ref{fig:spottedPredSE}). Adults of both
sexes spent considerable time in April, May and June hauled out. Adult spotted
seal males had a more protracted haul-out season compared to females, and more
time out of the water in June (Figure \ref{fig:spottedHOCal}). This likely
reflects the triad behavior in spotted seals when suitor males haul out with
nursing females. As with ribbon seals, the young-of-the-year records began after
weaning and the model predictions reflected development of in-water activities
(e.g.~swimming, foraging) in May and haul-out behavior centered around solar
noon.

@@

The second set of models, which included annual variation in haul-out
patterns, uncovered significant contributions for linear and quadratic
interactions between day and year for only spotted seals (\emph{day:year},
\(F_{15,115144}\)
= 4.445; \(p\) =
\textless0.001; \emph{day\textsuperscript{2}:year},
\(F_{15,115144}\)
= 5.854; \(p\) =
\textless0.001). Ribbon seals showed no significant
contribution for interactions between day and year (\emph{day:year},
\(F_{10,99510}\)
= 0.516; \(p\) =
0.880; \emph{day\textsuperscript{2}:year},
\(F_{10,99510}\)
= 0.549; \(p\) =
0.856). Predicted distributions of
haul-out activity were largely unimodal, but varied some among and
within years with respect to the timing and magnitude of haul-out peaks
(Figure \ref{fig:annualHO}). It is important to note that predicted
variation in annual haul-out patterns likely reflected both process error
and sampling variability. While we did remove any years where only one
deployment in a species + age:sex group was present, there were still
some years where the pattern shown was informed by a small number of
individuals that may not represent population-level patterns.

@@

The annual timing of peak haul-out probability for ribbon seals and adult
male spotted seals appeared to have
no relationship with the amount of yearly maximum sea-ice extent in the
study area. For ribbon seals and adult male spotted seals,
\(p\)-values were substantially larger than 0.05 (ribbon seal adult
females; \(R^{2}\) = 0.004, \(p\) = 0.896;
ribbon seal adult males: \(R^{2}\) = 0.059, \(p\) =
0.693; ribbon seal subadults: \(R^{2}\) =
0.007, \(p\) = 0.828; spotted seals
adult males: \(R^{2}\) = 0.004, \(p\) =
0.889). Adult female and subadult spotted seals both
showed a negative trend (peak haul-out occurs later in years with less
sea ice) but neither with a significant relationship\\
(spotted seal adult female: \(R^{2}\) =
0.456, \(p\) = 0.141; spotted seal
subadults: \(R^{2}\) = 0.369, \(p\) =
0.062).

@@

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g.~Udevitz et al.
(2009)) and a few select studies of ice-associated seals
(Perry, Stenson \& Buren, 2017; Hamilton \& Evans, 2018). In our study, ribbon seal haul-out behavior was notably
influenced by weather, with wind, temperature, and barometric pressure all being
important components of the model. Spotted seals were most affected by wind and
barometric pressure. For bearded seals, the model indicated wind and temperature
had the greatest impact. In general, and as might be expected, seals were more
likely to haul-out when daily temperatures were warmer, winds speeds were lower,
barometric pressure was higher, and precipitation was lower. Those weather
conditions are general indicators of increased solar radiation and lower
convective heat loss, both of which provide energetic benefits (see additional
discussion in Supplemental Material \hyperref[exploring-insolation-solar-radiation-as-a-model-covariate]{Exploring Insolation (Solar Radiation) as a
Model Covariate} regarding the potential use of solar radiation directly). Low
winds and precipitation could also enhance predator detection. Our results
highlight the importance of incorporating weather covariates when analyzing
haul-out behavior and calculating availability corrections for aerial surveys.

@@

We recognize that the species and ecosystems we studied are within the ancestral
and present-day environs of the Inupiat and Yup'ik people who, through many
uncredited contributions of traditional knowledge, provided early western
naturalists and scientists with much of what gets described as the `basic
biology' of Arctic seals. The deployment of bio-logging devices used in this
study were often done in collaboration with Alaska Native seal hunters and wiht the
approval of
their communities. We would like to especially acknowledge the Alaska communities of
Kotzebue, Koyuk, Nome, Nuiqsut, Scammon Bay, St.~Michael, Utqiaġvik, and Ulguniq
(Wainwright) and the following individuals: James Adams, Jeff Barger, David
Barr, Wendell Booth, Cyrus Harris, Nereus `Doc' Harris, Grover Harris, Lee
Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin, John Goodwin,
Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil Naylor Jr.,
Virgil Naylor Sr., Dan Savetilik, Chuck Schaeffer, Ross Schaeffer, Allen Stone,
and Randy Toshavik from Kotzebue; Merlin Henry from Koyuk; Tom
Gray from Nome; Vernon Long and Richard Tukle from Nuiqsuit; Morgan
Simon, River Simon, and Al Smith from Scammon Bay; Alex Niksik Jr.~from
St.~Michael; Billy Adams, James Aiken, Tim Aiken, Howard Kittick,
Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Bobby Sarren, and Joe Skin from Utqiaġvik;
Mary Ellen Ahmaogak, Enoch Oktollik, Shawn Oktollik, Stacey Osborn, and
Fred Rexford from Ulguniq. We would also like to acknowledge the support of the
Ice Seal Committee.

We are grateful for the assistance in catching and sampling seals by Ryan Adam,
James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Anna Bryan,
Vladamir Burkanov, Cynthia Christman, Sarah Coburn, Shawn Dahle, Rob Delong,
Stacy DiRocco, Deb Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Craig
George, Tracey Goldstein, Jeff Harris, Jason Herreman, Markus Horning, John
Jansen, Shawn Johnson, Charles Littnan, Lloyd Lowry, Brett McClintock, Erin
Moreland, Aaron Morris, Mark Nelson, Justin Olnes, Lorrie Rea, Bob Shears, Gay
Sheffield, Kayla Scheimreif, Brent Stewart, Alexy Trukhin, Dave Withrow, and
Heather Ziel. We also appreciate the commitment to science and safety by all
officers and crew of the NOAA ship \emph{Oscar Dyson}, the NOAA ship \emph{MacArthur II},
the \emph{MV Tayfun}, and the RV \emph{Thomas G. Thompson}.

Telemetry data from the Alaska Department of Fish and Game (ADF\&G) and the North
Slope Borough Department of Wildlife Management (NSB) were important
contributions to the findings presented here. Deployments in the western Bering
Sea were done in collaboration with Russian colleagues and North Pacific
Wildlife Consulting, LLC.
